---
title: "Data Science: Visualization"
author: "Gustavo Roa"
date: "2023-04-19"
output:
  html_document: default
---

```{r setup, include=FALSE}

# load the dataset
library(dslabs)
data(heights)

```

# Section 1 
## Describe Heights to ET
```{r} 
# load the dataset
library(dslabs)
data(heights)

# make a table of category proportions
prop.table(table(heights$sex))

my_data <- data(heights)

```

## Cumulative Distribution Function
```{r}
#a <- seq(min(my_data), max(my_data), length = 100)    # define range of values spanning the dataset
#cdf_function <- function(x) {    # computes prob. for a single value
    #mean(my_data <= x)
#}
#cdf_values <- sapply(a, cdf_function)
#plot(a, cdf_values)

```

## Normal distribution
```{r}

# define x as vector of male heights
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male"
x <- heights$height[index]

# calculate the mean and standard deviation manually
average <- sum(x)/length(x)
SD <- sqrt(sum((x - average)^2)/length(x))

# built-in mean and sd functions - note that the audio and printed values disagree
average <- mean(x)
SD <- sd(x)
c(average = average, SD = SD)

# calculate standard units
z <- scale(x)
#  2 standard deviations
mean(abs(z) < 2)

# calculate proportion of values within 2 SD of mean
mean(abs(z) < 2)

```
## The Normal CDF and pnorm
```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)
#We can estimate the probability that a male is taller than 70.5 inches with:

1 - pnorm(70.5, mean(x), sd(x))
#Code: Discretization and the normal approximation
# plot distribution of exact heights in data
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer
mean(x <= 68.5) - mean(x <= 67.5)
mean(x <= 69.5) - mean(x <= 68.5)
mean(x <= 70.5) - mean(x <= 69.5)

# probabilities in normal approximation match well
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))
pnorm(70.5, mean(x), sd(x)) - pnorm(69.5, mean(x), sd(x))

# probabilities in actual data over other ranges don't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))

```

## Definition of quantiles

```{r}
#Examples
#Load the heights dataset from the dslabs package:

library(dslabs)
data(heights)
#Use summary() on the heights$height variable to find the quartiles:

summary(heights$height)
#Find the percentiles of heights$height:

p <- seq(0.01, 0.99, 0.01)
percentiles <- quantile(heights$height, p)
#Confirm that the 25th and 75th percentiles match the 1st and 3rd quartiles. Note that quantile() returns a named vector. You can access the 25th and 75th percentiles like this (adapt the code for other percentile values):

percentiles[names(percentiles) == "25%"]
percentiles[names(percentiles) == "75%"]

#Theoretical quantiles
p <- seq(0.01, 0.99, 0.01)
theoretical_quantiles <- qnorm(p, 69, 3)

```

## Quantile-Quantile Plots
```{r}
# define x and z
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male"
x <- heights$height[index]
z <- scale(x)

# proportion of data below 69.5
mean(x <= 69.5)

# calculate observed and theoretical quantiles
p <- seq(0.05, 0.95, 0.05)
observed_quantiles <- quantile(x, p)
theoretical_quantiles <- qnorm(p, mean = mean(x), sd = sd(x))

# make QQ-plot
plot(theoretical_quantiles, observed_quantiles)
abline(0,1)

# make QQ-plot with scaled values
observed_quantiles <- quantile(z, p)
theoretical_quantiles <- qnorm(p)
plot(theoretical_quantiles, observed_quantiles)
abline(0,1)
```
# Section 2

## Creating a New Plot
```{r}

library(tidyverse)
library(dslabs)
data(murders)

ggplot(data = murders)

murders %>% ggplot()

p <- ggplot(data = murders)
class(p)
print(p)    # this is equivalent to simply typing p
p

```

## Layers
```{r}
#Code: Adding layers to a plot
library(tidyverse)
library(dslabs)
data(murders)

murders %>% ggplot() +
    geom_point(aes(x = population/10^6, y = total))
    
# add points layer to predefined ggplot object
p <- ggplot(data = murders)
p + geom_point(aes(population/10^6, total))

# add text layer to scatterplot
p + geom_point(aes(population/10^6, total)) +
    geom_text(aes(population/10^6, total, label = abb))

#Code: Example of aes behavior
# no error from this call
p_test <- p + geom_text(aes(population/10^6, total, label = abb))

# error - "abb" is not a globally defined variable and cannot be found outside of aes
#p_test <- p + geom_text(aes(population/10^6, total), label = abb) 

```

## Tinkering
```{r}
# change the size of the points
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb))
    
# move text labels slightly to the right
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
    
# simplify code by adding global aesthetic
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
p + geom_point(size = 3) +
    geom_text(nudge_x = 1.5)
    
# local aesthetics override global aesthetics
p + geom_point(size = 3) +
    geom_text(aes(x = 10, y = 800, label = "Hello there!"))

```
## Scales, Labels, and Colors
```{r}
##Code: Log-scale the x- and y-axis
# define p
library(tidyverse)
library(dslabs)
data(murders)
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))

# log base 10 scale the x-axis and y-axis
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.05) +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10")
    
# efficient log scaling of the axes
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10()

##Code: Add labels and title
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")

##Code: Change color of the points
# redefine p to be everything except the points layer
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
    
# make all points blue
p + geom_point(size = 3, color = "blue")

# color points by region
p + geom_point(aes(col = region), size = 3)

##Code: Add a line with average murder rate
# define average murder rate
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    pull(rate)
    
# basic line with average murder rate for the country
p <- p + geom_point(aes(col = region), size = 3) +
    geom_abline(intercept = log10(r))    # slope is default of 1
    
# change line to dashed and dark grey, line under points
p + 
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3)

##Code: Change legend title
p <- p + scale_color_discrete(name = "Region")    # capitalize legend title

```
## Add-on Packages
```{r}
##Code: Adding themes
# theme used for graphs in the textbook and course
library(dslabs)
ds_theme_set()

# themes from ggthemes
library(ggthemes)
p + theme_economist()    # style of the Economist magazine
p + theme_fivethirtyeight()    # style of the FiveThirtyEight website

##Code: Putting it all together to assemble the plot
# load libraries
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(dslabs)
data(murders)

# define the intercept
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    .$rate
    
# make the plot, combining all elements
murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) +
    geom_text_repel() +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010") +
    scale_color_discrete(name = "Region") +
    theme_economist()

```

## Other Examples
```{r}
##Code: Histograms in ggplot2
# load heights data
library(tidyverse)
library(dslabs)
data(heights)

# define p
p <- heights %>%
    filter(sex == "Male") %>%
    ggplot(aes(x = height))
    
# basic histograms
p + geom_histogram()
p + geom_histogram(binwidth = 1)

# histogram with blue fill, black outline, labels and title
p + geom_histogram(binwidth = 1, fill = "blue", col = "black") +
    xlab("Male heights in inches") +
    ggtitle("Histogram")

##Code: Smooth density plots in ggplot2
p + geom_density()
p + geom_density(fill = "blue")

##Code: Quantile-quantile plots in ggplot2
# basic QQ-plot
p <- heights %>% filter(sex == "Male") %>%
    ggplot(aes(sample = height))
p + geom_qq()

# QQ-plot against a normal distribution with same mean/sd as data
params <- heights %>%
    filter(sex == "Male") %>%
    summarize(mean = mean(height), sd = sd(height))
    p + geom_qq(dparams = params) +
    geom_abline()
    
# QQ-plot of scaled data against the standard normal distribution
heights %>%
    ggplot(aes(sample = scale(height))) +
    geom_qq() +
    geom_abline()

##Code: Grids of plots with the gridExtra package
# define plots p1, p2, p3
p <- heights %>% filter(sex == "Male") %>% ggplot(aes(x = height))
p1 <- p + geom_histogram(binwidth = 1, fill = "blue", col = "black")
p2 <- p + geom_histogram(binwidth = 2, fill = "blue", col = "black")
p3 <- p + geom_histogram(binwidth = 3, fill = "blue", col = "black")

# arrange plots next to each other in 1 row, 3 columns
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)

```
# Section 3

## Gapminder Dataset
```{r}
#Code
# load and inspect gapminder data
library(dslabs)
data(gapminder)
head(gapminder)

# compare infant mortality in Sri Lanka and Turkey
gapminder %>% 
    filter(year == 2015 & country %in% c("Sri Lanka", "Turkey")) %>% 
    select(country, infant_mortality)

```

## Life Expectancy and Fertility Rates
```{r}
# basic scatterplot of life expectancy versus fertility
ds_theme_set()    # set plot theme
filter(gapminder, year == 1962) %>%
    ggplot(aes(fertility, life_expectancy)) +
    geom_point()
    
# add color as continent
filter(gapminder, year == 1962) %>%
    ggplot(aes(fertility, life_expectancy, color = continent)) +
    geom_point()
```

## Faceting
```{r}
# facet by continent and year
filter(gapminder, year %in% c(1962, 2012)) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_grid(continent ~ year)

# facet by year only
filter(gapminder, year %in% c(1962, 2012)) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_grid(. ~ year)

# facet by year, plots wrapped onto multiple rows
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>%
    filter(year %in% years & continent %in% continents) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_wrap(~year)
```

## Time Series Plots
```{r}
#Code: Single time series
# scatterplot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_point()

# line plot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_line()

#Code: Multiple time series
# line plot fertility time series for two countries- only one line (incorrect)
countries <- c("South Korea", "Germany")
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility)) +
    geom_line()
    
# line plot fertility time series for two countries - one line per country
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, group = country)) +
    geom_line()

# fertility time series for two countries - lines colored by country
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, col = country)) +
    geom_line()

#Code: Adding text labels to a plot
# life expectancy time series - lines colored by country and labeled, no legend
labels <- data.frame(country = countries, x = c(1975, 1965), y = c(60, 72))
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, life_expectancy, col = country)) +
    geom_line() +
    geom_text(data = labels, aes(x, y, label = country), size = 5) +
    theme(legend.position = "none")
```

## Transformations
```{r}
# add dollars per day variable
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)

# histogram of dollars per day
past_year <- 1970
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black")

# repeat histogram with log2 scaled data
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(log2(dollars_per_day))) +
    geom_histogram(binwidth = 1, color = "black")
    
# repeat histogram with log2 scaled x-axis
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2")

```

## Stratify and Boxplot
```{r}
#Code: Boxplot of GDP by region
# add dollars per day variable
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
    
# number of regions
length(levels(gapminder$region))

# boxplot of GDP by region in 1970
past_year <- 1970
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(region, dollars_per_day))
p + geom_boxplot()

# rotate names on x-axis
p + geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Code: The reorder function
# by default, factor order is alphabetical
fac <- factor(c("Asia", "Asia", "West", "West", "West"))
levels(fac)

# reorder factor by the category means
value <- c(10, 11, 12, 6, 4)
fac <- reorder(fac, value, FUN = mean)
levels(fac)

#Code: Enhanced boxplot ordered by median income, scaled, and showing data
# reorder by median income and color by continent
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%    # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("")
p

# log2 scale y-axis
p + scale_y_continuous(trans = "log2")

# add data points
p + scale_y_continuous(trans = "log2") + geom_point(show.legend = FALSE)
```
## Comparing Distributions
```{r}

#Code: Histogram of income in West versus developing world, 1970 and 2010
# add dollars per day variable and define past year
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
past_year <- 1970

# define Western countries
west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")

# facet by West vs devloping
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(. ~ group)

# facet by West/developing and year
present_year <- 2010
gapminder %>%
    filter(year %in% c(past_year, present_year) & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)

#Code: Income distribution of West versus developing world, only countries with data 
# define countries that have data available in both years
country_list_1 <- gapminder %>%
    filter(year == past_year & !is.na(dollars_per_day)) %>% .$country
    country_list_2 <- gapminder %>%
    filter(year == present_year & !is.na(dollars_per_day)) %>% .$country
    country_list <- intersect(country_list_1, country_list_2)

# make histogram including only countries with data available in both years
gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%    # keep only selected countries
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)

#Code: Boxplots of income in West versus developing world, 1970 and 2010
p <- gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
    ggplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("") + scale_y_continuous(trans = "log2")
    
 p + geom_boxplot(aes(region, dollars_per_day, fill = continent)) +
     facet_grid(year ~ .)
 
 # arrange matching boxplots next to each other, colored by year
 p + geom_boxplot(aes(region, dollars_per_day, fill = factor(year)))
 
```

## Density Plots
```{r}
#Code: Faceted smooth density plots
# see the code below the previous video for variable definitions
# smooth density plots - area under each curve adds to 1
gapminder %>%
    filter(year == past_year & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>% group_by(group) %>%
    summarize(n = n()) %>% knitr::kable()

# smooth density plots - variable counts on y-axis
p <- gapminder %>%
    filter(year == past_year & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
    scale_x_continuous(trans = "log2")
p + geom_density(alpha = 0.2, bw = 0.75) + facet_grid(year ~ .)

#Code: Add new region groups with case_when
# add group as a factor, grouping regions
gapminder <- gapminder %>%
    mutate(group = case_when(
            .$region %in% west ~ "West",
            .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
            .$region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
            .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
            TRUE ~ "Others"))

# reorder factor levels
gapminder <- gapminder %>%
    mutate(group = factor(group, levels = c("Others", "Latin America", "East Asia", "Sub-Saharan Africa", "West")))

#Code: Stacked density plot
# note you must redefine p with the new gapminder object first
p <- gapminder %>%
  filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    ggplot(aes(dollars_per_day, fill = group)) +
    scale_x_continuous(trans = "log2")

# stacked density plot
p + geom_density(alpha = 0.2, bw = 0.75, position = "stack") +
    facet_grid(year ~ .)

#Code: Weighted stacked density plot
# weighted stacked density plot
gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    group_by(year) %>%
    mutate(weight = population/sum(population*2)) %>%
    ungroup() %>%
    ggplot(aes(dollars_per_day, fill = group, weight = weight)) +
    scale_x_continuous(trans = "log2") +
    geom_density(alpha = 0.2, bw = 0.75, position = "stack") + facet_grid(year ~ .)

```

## Ecological Fallacy
```{r}
# define gapminder
library(tidyverse)
library(dslabs)
data(gapminder)

# add additional cases
gapminder <- gapminder %>%
    mutate(group = case_when(
        .$region %in% west ~ "The West",
        .$region %in% "Northern Africa" ~ "Northern Africa",
        .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
        .$region == "Southern Asia" ~ "Southern Asia",
        .$region %in% c("Central America", "South America", "Caribbean") ~ "Latin America",
        .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
        .$region %in% c("Melanesia", "Micronesia", "Polynesia") ~ "Pacific Islands"))

# define a data frame with group average income and average infant survival rate
surv_income <- gapminder %>%
    filter(year %in% present_year & !is.na(gdp) & !is.na(infant_mortality) & !is.na(group)) %>%
    group_by(group) %>%
    summarize(income = sum(gdp)/sum(population)/365,
                        infant_survival_rate = 1 - sum(infant_mortality/1000*population)/sum(population))
surv_income %>% arrange(income)

# plot infant survival versus income, with transformed axes
surv_income %>% ggplot(aes(income, infant_survival_rate, label = group, color = group)) +
    scale_x_continuous(trans = "log2", limit = c(0.25, 150)) +
    scale_y_continuous(trans = "logit", limit = c(0.875, .9981),
                                       breaks = c(.85, .90, .95, .99, .995, .998)) +
    geom_label(size = 3, show.legend = FALSE)

```
# Section 4

## Show the Data
```{r}

# dot plot showing the data
heights %>% ggplot(aes(sex, height)) + geom_point()

# jittered, alpha blended point plot
heights %>% ggplot(aes(sex, height)) + geom_jitter(width = 0.1, alpha = 0.2)
```

## Ease Comparisons: Compared Visual Cues Should Be Adjacent
```{r}

color_blind_friendly_cols <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

p1 <- data.frame(x = 1:8, y = 1:8, col = as.character(1:8)) %>%
    ggplot(aes(x, y, color = col)) +
    geom_point(size = 5)
p1 + scale_color_manual(values = color_blind_friendly_cols)

```

## Slope Charts
```{r}

library(tidyverse)
library(dslabs)
data(gapminder)

west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")

dat <- gapminder %>%
    filter(year %in% c(2010, 2015) & region %in% west & !is.na(life_expectancy) & population > 10^7)

dat %>%
    mutate(location = ifelse(year == 2010, 1, 2),
           location = ifelse(year == 2015 & country %in% c("United Kingdom", "Portugal"),
                             location + 0.22, location),
           hjust = ifelse(year == 2010, 1, 0)) %>%
    mutate(year = as.factor(year)) %>%
    ggplot(aes(year, life_expectancy, group = country)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    geom_text(aes(x = location, label = country, hjust = hjust), show.legend = FALSE) +
    xlab("") +
    ylab("Life Expectancy") 

#Code: Bland-Altman plot
library(ggrepel)
dat %>%
    mutate(year = paste0("life_expectancy_", year)) %>%
    select(country, year, life_expectancy) %>% spread(year, life_expectancy) %>%
    mutate(average = (life_expectancy_2015 + life_expectancy_2010)/2,
                difference = life_expectancy_2015 - life_expectancy_2010) %>%
    ggplot(aes(average, difference, label = country)) +
    geom_point() +
    geom_text_repel() +
    geom_abline(lty = 2) +
    xlab("Average of 2010 and 2015") +
    ylab("Difference between 2015 and 2010")

```

## Case Study: Vaccines
```{r}
#Code: Tile plot of measles rate by year and state
# import data and inspect
library(tidyverse)
library(dslabs)
data(us_contagious_diseases)
str(us_contagious_diseases)

# assign dat to the per 10,000 rate of measles, removing Alaska and Hawaii and adjusting for weeks reporting
the_disease <- "Measles"
dat <- us_contagious_diseases %>%
    filter(!state %in% c("Hawaii", "Alaska") & disease == the_disease) %>%
    mutate(rate = count / population * 10000 * 52/weeks_reporting) %>%
    mutate(state = reorder(state, rate))

# plot disease rates per year in California
dat %>% filter(state == "California" & !is.na(rate)) %>%
    ggplot(aes(year, rate)) +
    geom_line() +
    ylab("Cases per 10,000") +
    geom_vline(xintercept=1963, col = "blue")

# tile plot of disease rate by state and year
dat %>% ggplot(aes(year, state, fill=rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "Reds"), trans = "sqrt") +
    geom_vline(xintercept = 1963, col = "blue") +
    theme_minimal() + theme(panel.grid = element_blank()) +
    ggtitle(the_disease) +
    ylab("") +
    xlab("")

#Code: Line plot of measles rate by year and state
# compute US average measles rate by year
avg <- us_contagious_diseases %>%
    filter(disease == the_disease) %>% group_by(year) %>%
    summarize(us_rate = sum(count, na.rm = TRUE)/sum(population, na.rm = TRUE)*10000)

# make line plot of measles rate by year by state
dat %>%
    filter(!is.na(rate)) %>%
    ggplot() +
    geom_line(aes(year, rate, group = state), color = "grey50", 
        show.legend = FALSE, alpha = 0.2, size = 1) +
    geom_line(mapping = aes(year, us_rate), data = avg, size = 1, col = "black") +
    scale_y_continuous(trans = "sqrt", breaks = c(5, 25, 125, 300)) +
    ggtitle("Cases per 10,000 by state") +
    xlab("") +
    ylab("") +
    geom_text(data = data.frame(x = 1955, y = 50),
        mapping = aes(x, y, label = "US average"), color = "black") +
    geom_vline(xintercept = 1963, col = "blue")

```

## Titanic Survival Exercises
```{r} 
#The Titanic was a British ocean liner that struck an iceberg and sunk on its maiden voyage in 1912 from the United Kingdom to New York. More than 1,500 of the estimated 2,224 passengers and crew died in the accident, making this one of the largest maritime disasters ever outside of war. The ship carried a wide range of passengers of all ages and both genders, from luxury travelers in first-class to immigrants in the lower classes. However, not all passengers were equally likely to survive the accident. We use real data about a selection of 891 passengers to learn who was on the Titanic and which passengers were more likely to survive.

options(digits = 3)    
library(tidyverse)
library(titanic)
titanic <- titanic_train %>%
  select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare) %>%
  mutate(Survived = factor(Survived),
         Pclass = factor(Pclass),
         Sex = factor(Sex))
  

#Question 1: Variable Types

#Question 2: Demographics of Titanic Passengers
total <- titanic %>%
  filter(Sex %in% c("female","male")) %>%
  count(Sex)
total

age40 <- titanic %>%
  filter(Age == 40 & Sex %in% c("female","male")) %>%
  count(Sex)
age40

max_age <- titanic %>%
  filter(!is.na(Age)) %>%
  group_by(Sex) %>%
  summarize(max = max(Age))
max_age

titanic %>%
  filter(!is.na(Age)) %>%
  ggplot(aes(x = Age, group = Sex, fill = Sex)) + 
  geom_density(alpha = 0.2, bw = 10) 

#Question 3: QQ-plot of Age Distribution

params <- titanic %>%
    filter(!is.na(Age)) %>%
    summarize(mean = mean(Age), sd = sd(Age))

p <- titanic %>%
  ggplot(aes(sample = Age)) +
  geom_qq(dparams = params) +
  geom_abline()
p

#Question 4: Survival by Sex
p <- titanic %>%
  ggplot(aes(x = Survived, fill = Sex)) +
  geom_bar(position = position_dodge())
p

#Question 5: Survival by Age
p <- titanic %>%
  ggplot(aes(x = Age, y = ..count.., fill = Survived)) +
  geom_density(alpha = 0.2)
p

#Question 6: Survival by Fare
p <- titanic %>%
  filter(Fare > 0) %>%
  ggplot(aes(Survived, Fare)) +
  geom_boxplot(alpha = 0.2) +
  scale_y_continuous(trans = "log2") +
  geom_point(show.legend = FALSE) + 
  geom_jitter()
p 

#Question 7: Survival by Passenger Class
p <- titanic %>%
  ggplot(aes(x = Pclass, fill = Survived)) +
  geom_bar()
p

p <- titanic %>%
  ggplot(aes(x = Pclass, fill = Survived)) +
  geom_bar(position = position_fill())
p

p <- titanic %>%
  ggplot(aes(x = Survived, fill = Pclass)) +
  geom_bar(position = position_fill())
p

#Question 8: Survival by Age, Sex and Passenger Class
p <- titanic %>%
  ggplot(aes(x = Age, y = ..count.., fill = Survived)) +
  geom_density(alpha = 0.2) +
  facet_grid(Pclass ~ Sex)
p



```

## Properties of Stars Exercises
```{r}
#Background
#Astronomy is one of the oldest data-driven sciences. In the late 1800s, the director of the Harvard College Observatory hired women to analyze astronomical data, which at the time was done using photographic glass plates. These women became known as the “Harvard Computers”. They computed the position and luminosity of various astronomical objects such as stars and galaxies. (If you are interested, you can learn more about the Harvard Computers External link). Today, astronomy is even more of a data-driven science, with an inordinate amount of data being produced by modern instruments every day.

#In the following exercises we will analyze some actual astronomical data to inspect properties of stars, their absolute magnitude (which relates to a star's luminosity, or brightness), temperature and type (spectral class).

#Libraries and Options

library(tidyverse)
library(dslabs)
data(stars)
options(digits = 3)   # report 3 significant digits

#Question 1
#a. What is the mean magnitude?
mean(stars$magnitude)
#b. What is the standard deviation of magnitude?
sd(stars$magnitude)

#Question 2
stars %>%
  ggplot(aes(magnitude)) +
  geom_density()

#Question 3
stars %>%
  ggplot(aes(temp)) +
  geom_density()

#Question 4
stars %>%
  ggplot(aes(temp, magnitude)) +
  geom_point()

#Question 5
stars %>%
  ggplot(aes(x=log10(temp), magnitude)) +
  scale_x_reverse() +
  scale_y_reverse() +
  geom_point()

#Question 6
#These stars are in the lower left of the plot from question 5. There are 4 stars in this region.

#Question 7
#Giants are in the upper right corner of the plot and generally have temperatures below 6000K.

#Question 8
stars %>%
    ggplot(aes(log10(temp), magnitude)) +
    geom_point() +
    geom_text(aes(label = star)) +
    scale_x_reverse() +
    scale_y_reverse()

#Question 9
stars %>%
    ggplot(aes(log10(temp), magnitude, col = type)) +
    geom_point() +
    scale_x_reverse() +
    scale_y_reverse()


```

## Climate Change Exercises
```{r}
#Libraries and Data Import
data(temp_carbon)
data(greenhouse_gases)
data(historic_co2)

#Question 1
#a 
temp_carbon %>%
    .$year %>%
    max()
#b
temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    pull(year) %>%
    max()
#c
#temp_carbon %>%
    #filter(!is.na(carbon_emissions)) %>%
    #max(year)
#d
temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    .$year %>%
    max()
#e
temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    select(year) %>%
    max()
#f
temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    max(.$year)

#Question 2
#What is the first year for which carbon emissions (carbon_emissions) data are available?

temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    .$year %>%
    min()

#What is the last year for which carbon emissions data are available?

temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    .$year %>%
    max()

#How many times larger were carbon emissions in the last year relative to the first year?

carbon1 <- temp_carbon %>%
    filter(year == 1751) %>%
    .$carbon_emissions

carbon2 <- temp_carbon %>%
    filter(year == 2014) %>%
    .$carbon_emissions

carbon2/carbon1

#Question 3
#What is the first year for which global temperature anomaly (temp_anomaly) data are available?

temp_carbon %>%
    filter(!is.na(temp_anomaly)) %>%
    .$year %>%
    min()

#What is the last year for which global temperature anomaly data are available?

temp_carbon %>%
    filter(!is.na(temp_anomaly)) %>%
    .$year %>%
    max()

#How many degrees Celsius has temperature increased over the date range? Compare the temperatures in the most recent year versus the oldest year.

temp1 <- temp_carbon %>%
    filter(year == "1880") %>%
    .$temp_anomaly

temp2 <- temp_carbon %>%
    filter(year == "2018") %>%
    .$temp_anomaly

temp2 - temp1

#Question 4

 p <- p + geom_hline(aes(yintercept = 0), col = "blue")
 
#Question 5
 
 p <- temp_carbon %>%
  filter(!is.na(temp_anomaly)) %>%
  ggplot(aes(year, temp_anomaly)) +
  geom_line() + 
  geom_hline(aes(yintercept=0), col='blue') +
  ylab("Temperature anomaly (degrees C)") +
  ggtitle("Temperature anomaly relative to 20th century mean, 1880-2018") +
  geom_text(aes(x = 2000, y = 0.05, label="20th century mean"), col='blue')
p

#Question 6

#When was the earliest year with a temperature above the 20th century mean? 1940

#When was the last year with an average temperature below the 20th century mean? 1976

#In what year did the temperature anomaly exceed 0.5 degrees Celsius for the first time? 1997

#Question 7

temp_carbon %>%
    filter(!is.na(temp_anomaly)) %>%
    ggplot(aes(year, temp_anomaly)) +
    geom_line() +
    geom_line(aes(year, land_anomaly), col = "red") +
    geom_line(aes(year, ocean_anomaly), col = "blue") +
    ylab("Temperature anomaly (degrees C)") +
    xlim(c(1880, 2018)) +
    ggtitle("Temperature anomaly on land and ocean")

#Question 8

greenhouse_gases %>%
    ggplot(aes(year, concentration)) +
    geom_line() +
    facet_grid(gas ~ ., scales = "free") +
    geom_vline(xintercept = 1850, col='red') +
    ylab("Concentration (ch4/n2o ppb, co2 ppm)") +
    ggtitle("Atmospheric greenhouse gas concentration by year, 0-2000")

#Question 9

#Which gas was stable at approximately 275 ppm/ppb until around 1850? co2

#Which gas more than doubled in concentration since 1850? ch4

#Which gas decreased in concentration since 1850? none

#Which gas had the smallest magnitude change since 1850? n2o

#Which gas increased exponentially in concentration after 1850? all

#Question 10

temp_carbon %>%
    filter(!is.na(carbon_emissions)) %>%
    ggplot(aes(year, carbon_emissions)) +
    geom_line() +
    ylab("Carbon emissions (metric tons)") +
    ggtitle("Annual global carbon emissions, 1751-2014")

#Question 11

co2_time <- historic_co2 %>%
  filter(!is.na(co2)) %>%
  ggplot(aes(year, co2, col=source)) +
  geom_line() +
  ggtitle("Atmospheric CO2 concentration, -800,000 BC to today") +
  ylab("co2 (ppmv)")
co2_time

#Question 12
#a
co2_time <- historic_co2 %>%
    ggplot(aes(year, co2, col = source)) +
    geom_line() +
    ggtitle("Atmospheric CO2 concentration, -800,000 BC to today") +
    ylab("co2 (ppmv)")
co2_time + xlim(-800000, -775000)
#b
co2_time <- historic_co2 %>%
    ggplot(aes(year, co2, col = source)) +
    geom_line() +
    ggtitle("Atmospheric CO2 concentration, -800,000 BC to today") +
    ylab("co2 (ppmv)")
co2_time + xlim(-375000, -330000)
#c
co2_time <- historic_co2 %>%
    ggplot(aes(year, co2, col = source)) +
    geom_line() +
    ggtitle("Atmospheric CO2 concentration, -800,000 BC to today") +
    ylab("co2 (ppmv)")
co2_time + xlim(-140000, -120000)
#d
co2_time <- historic_co2 %>%
    ggplot(aes(year, co2, col = source)) +
    geom_line() +
    ggtitle("Atmospheric CO2 concentration, -800,000 BC to today") +
    ylab("co2 (ppmv)")
co2_time + xlim(-3000, 2018)



```

