---
title: "Probability"
author: "Gustavo Roa"
date: "2023-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1 
## Monte Carlo Simulations
```{r}
beads <- rep(c("red", "blue"), times = c(2,3))    # create an urn with 2 red, 3 blue
beads    # view beads object
sample(beads, 1)    # sample 1 bead at random

B <- 10000    # number of times to draw 1 bead
events <- replicate(B, sample(beads, 1))    # draw 1 bead, B times
tab <- table(events)    # make a table of outcome counts
tab    # view count table
prop.table(tab)    # view table of outcome proportions

# no replacement
sample(beads, 5) 
sample(beads, 5)
sample(beads, 5)

#with replacement using sample 
events <- sample(beads,B, replace = TRUE)    
prop.table(table(events))    


```
## Setting the Random Seed
```{r}
set.seed(1986) 

#Important note on seeds in R 3.5 versus R 3.6 and later
set.seed(1)
set.seed(1, sample.kind="Rounding")    # will make R 3.6 generate a seed as in R 3.5

```

## Using the mean Function for Probability
```{r}

beads <- rep(c("red", "blue"), times = c(2,3))
beads

mean(beads == "blue")


```

## Assessment: Introduction to Discrete Probability
NA

## Combinations and Permutations
```{r}
#Code: Introducing paste() and expand.grid()
# joining strings with paste
number <- "Three"
suit <- "Hearts"
paste(number, suit)

# joining vectors element-wise with paste
paste(letters[1:5], as.character(1:5))

# generating combinations of 2 vectors with expand.grid
expand.grid(pants = c("blue", "black"), shirt = c("white", "grey", "plaid"))

#Code: Generating a deck of cards
suits <- c("Diamonds", "Clubs", "Hearts", "Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nine", "Ten", "Jack", "Queen", "King")
deck <- expand.grid(number = numbers, suit = suits)
deck <- paste(deck$number, deck$suit)

# probability of drawing a king
kings <- paste("King", suits)
mean(deck %in% kings)

#Code: Permutations and combinations
#Correction: The code shown does not generate all 7 digit phone numbers because phone numbers can have repeated digits. It generates all possible 7 digit numbers without repeats.

library(gtools)
permutations(5,2)    # ways to choose 2 numbers in order from 1:5
all_phone_numbers <- permutations(10, 7, v = 0:9)
n <- nrow(all_phone_numbers)
index <- sample(n, 5)
all_phone_numbers[index,]

permutations(3,2)    # order matters
combinations(3,2)    # order does not matter

#Code: Probability of drawing a second king given that one king is drawn
hands <- permutations(52,2, v = deck)
first_card <- hands[,1]
second_card <- hands[,2]
sum(first_card %in% kings)

sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)

#Code: Probability of a natural 21 in blackjack
aces <- paste("Ace", suits)
facecard <- c("King", "Queen", "Jack", "Ten")
facecard <- expand.grid(number = facecard, suit = suits)
facecard <- paste(facecard$number, facecard$suit)

hands <- combinations(52, 2, v=deck) # all possible hands

# probability of a natural 21 given that the ace is listed first in `combinations`
mean(hands[,1] %in% aces & hands[,2] %in% facecard)

# probability of a natural 21 checking for both ace first and ace second
mean((hands[,1] %in% aces & hands[,2] %in% facecard)|(hands[,2] %in% aces & hands[,1] %in% facecard))

#Code: Monte Carlo simulation of natural 21 in blackjack
#Note that your exact values will differ because the process is random and the seed is not set.

# code for one hand of blackjack
hand <- sample(deck, 2)
hand

# code for B=10,000 hands of blackjack
B <- 10000
results <- replicate(B, {
    hand <- sample(deck, 2)
    (hand[1] %in% aces & hand[2] %in% facecard) | (hand[2] %in% aces & hand[1] %in% facecard)
})
mean(results)

```
## The Birthday Problem
```{r}
#Code: The birthday problem
# checking for duplicated bdays in one 50 person group
n <- 50
bdays <- sample(1:365, n, replace = TRUE)    # generate n random birthdays
any(duplicated(bdays))    # check if any birthdays are duplicated

# Monte Carlo simulation with B=10000 replicates
B <- 10000
results <- replicate(B, {    # returns vector of B logical values
    bdays <- sample(1:365, n, replace = TRUE)
    any(duplicated(bdays))
})
mean(results)    # calculates proportion of groups with duplicated bdays
```
## Sapply 
```{r}
#Code: Function for birthday problem Monte Carlo simulations
#Note that the function body of compute_prob() is the code that we wrote in the previous video. If we write this code as a function, we can use sapply() to apply this function to several values of n.

# function to calculate probability of shared bdays across n people
compute_prob <- function(n, B = 10000) {
	same_day <- replicate(B, {
    	bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}

n <- seq(1, 60)

#Code: Element-wise operation over vectors and sapply
x <- 1:10
sqrt(x)    # sqrt operates on each element of the vector

y <- 1:10
x*y    # * operates element-wise on both vectors

#compute_prob(n)    # does not iterate over the vector n without sapply

x <- 1:10
sapply(x, sqrt)    # this is equivalent to sqrt(x)

prob <- sapply(n, compute_prob)    # element-wise application of compute_prob to n
plot(n, prob)

#Code: Computing birthday problem probabilities with sapply
# function for computing exact probability of shared birthdays for any n
exact_prob <- function(n){
    prob_unique <- seq(365, 365-n+1)/365   # vector of fractions for mult. rule
    1 - prod(prob_unique)    # calculate prob of no shared birthdays and subtract from 1
}

# applying function element-wise to vector of n values
eprob <- sapply(n, exact_prob)

# plotting Monte Carlo results and exact probabilities on same graph
plot(n, prob)    # plot Monte Carlo results
lines(n, eprob, col = "red")    # add line for exact prob

```
## How Many Monte Carlo Experiments are Enough?
```{r}
#Code: Estimating a practical value of B
#This code runs Monte Carlo simulations to estimate the probability of shared birthdays using several B values and plots the results. When B is large enough that the estimated probability stays stable, then we have selected a useful value of B.

B <- 10^seq(1, 5, len = 100)    # defines vector of many B values
compute_prob <- function(B, n = 22){    # function to run Monte Carlo simulation with each B
	same_day <- replicate(B, {
    	bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}

prob <- sapply(B, compute_prob)    # apply compute_prob to many values of B
plot(log10(B), prob, type = "l")    # plot a line graph of estimates 

```
## The Monty Hall Problem
```{r}
#Code: Monte Carlo simulation of stick strategy
B <- 10000
stick <- replicate(B, {
	doors <- as.character(1:3)
	prize <- sample(c("car","goat","goat"))    # puts prizes in random order
	prize_door <- doors[prize == "car"]    # note which door has prize
	my_pick  <- sample(doors, 1)    # note which door is chosen
	show <- sample(doors[!doors %in% c(my_pick, prize_door)],1)    # open door with no prize that isn't chosen
	stick <- my_pick    # stick with original door
	stick == prize_door    # test whether the original door has the prize
})
mean(stick)    # probability of choosing prize door when sticking

#Code: Monte Carlo simulation of switch strategy
switch <- replicate(B, {
	doors <- as.character(1:3)
	prize <- sample(c("car","goat","goat"))    # puts prizes in random order
	prize_door <- doors[prize == "car"]    # note which door has prize
	my_pick  <- sample(doors, 1)    # note which door is chosen first
	show <- sample(doors[!doors %in% c(my_pick, prize_door)], 1)    # open door with no prize that isn't chosen
	switch <- doors[!doors%in%c(my_pick, show)]    # switch to the door that wasn't chosen first or opened
	switch == prize_door    # test whether the switched door has the prize
})
mean(switch)    # probability of choosing prize door when switching

```
## Assessment - Olympic Running
NA

## Questions 3 and 4: Esophageal cancer and alcohol/tobacco use, part 1
NA

## Questions 5 and 6: Esophageal cancer and alcohol/tobacco use, part 2
NA

# Section 2
## Continuous Probability
```{r}
library(tidyverse)
library(dslabs)
data(heights)

x <- heights %>% filter(sex=="Male") %>% pull(height)

#Given a vector x, we can define a function for computing the CDF of x using:

F <- function(a) mean(x <= a)

1 - F(70)    # probability of male taller than 70 inches

```
## Theoretical Distribution
```{r}
#Code: Using pnorm() to calculate probabilities
#Given male heights x:

library(tidyverse)
library(dslabs)
data(heights)

x <- heights %>% filter(sex=="Male") %>% pull(height)

#We can estimate the probability that a male is taller than 70.5 inches using:

1 - pnorm(70.5, mean(x), sd(x))

#Code: Discretization and the normal approximation
# plot distribution of exact heights in data
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer
mean(x <= 68.5) - mean(x <= 67.5)
mean(x <= 69.5) - mean(x <= 68.5)
mean(x <= 70.5) - mean(x <= 69.5)

# probabilities in normal approximation match well
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))
pnorm(70.5, mean(x), sd(x)) - pnorm(69.5, mean(x), sd(x))

# probabilities in actual data over other ranges don't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))

```
## Plotting the Probability Density
```{r}
library(tidyverse)
x <- seq(-4, 4, length = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x, f)) +
    geom_line()

#Note that dnorm() gives densities for the standard normal distribution by default. Probabilities for alternative normal distributions with mean mu and standard deviation sigma can be evaluated with:
  #dnorm(z, mu, sigma)

```
## Monte Carlo Simulations 
```{r}
#Code: Generating normally distributed random numbers
# define x as male heights from dslabs data
library(tidyverse)
library(dslabs)
data(heights)

x <- heights %>% filter(sex=="Male") %>% pull(height)
# generate simulated height data using normal distribution - both datasets should have n observations
n <- length(x)
avg <- mean(x)
s <- sd(x)
simulated_heights <- rnorm(n, avg, s)

# plot distribution of simulated_heights
data.frame(simulated_heights = simulated_heights) %>%
    ggplot(aes(simulated_heights)) +
    geom_histogram(color="black", binwidth = 2)

#Code: Monte Carlo simulation of tallest person over 7 feet
B <- 10000
tallest <- replicate(B, {
    simulated_data <- rnorm(800, avg, s)    # generate 800 normally distributed random heights
    max(simulated_data)    # determine the tallest height
})
mean(tallest >= 7*12)    # proportion of times that tallest person exceeded 7 feet (84 inches)

```
## Other Continuous Distributions
```{r}
x <- seq(-4, 4, length.out = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x,f)) +
    geom_line()
```

## Assessment - Continuous Probability
NA 

# Section 3 
## Random Variables
```{r}
#Code: Modeling a random variable
# define random variable x to be 1 if blue, 0 otherwise
beads <- rep(c("red", "blue"), times = c(2, 3))
x <- ifelse(sample(beads, 1) == "blue", 1, 0)

# demonstrate that the random variable is different every time
ifelse(sample(beads, 1) == "blue", 1, 0)
ifelse(sample(beads, 1) == "blue", 1, 0)
ifelse(sample(beads, 1) == "blue", 1, 0)
```
## Sampling Models
```{r}
#Monte Carlo simulation: Chance of casino losing money on roulette
#We build a sampling model for the random variable  that represents the casino's total winnings. 

# sampling model 1: define urn, then sample
color <- rep(c("Black", "Red", "Green"), c(18, 18, 2)) # define the urn for the sampling model
n <- 1000
X <- sample(ifelse(color == "Red", -1, 1), n, replace = TRUE)
X[1:10]

# sampling model 2: define urn inside sample function by noting probabilities
x <- sample(c(-1, 1), n, replace = TRUE, prob = c(9/19, 10/19))    # 1000 independent draws
S <- sum(x)    # total winnings = sum of draws
S

#We use the sampling model to run a Monte Carlo simulation and use the results to estimate the probability of the casino losing money.

n <- 1000    # number of roulette players
B <- 10000    # number of Monte Carlo experiments
S <- replicate(B, {
    X <- sample(c(-1,1), n, replace = TRUE, prob = c(9/19, 10/19))    # simulate 1000 spins
    sum(X)    # determine total profit
})

mean(S < 0)    # probability of the casino losing money

#We can plot a histogram of the observed values of S as well as the normal density curve based on the mean and standard deviation of S.
library(tidyverse)
s <- seq(min(S), max(S), length = 100)    # sequence of 100 values across range of S
normal_density <- data.frame(s = s, f = dnorm(s, mean(S), sd(S))) # generate normal density for S
data.frame (S = S) %>%    # make data frame of S for histogram
    ggplot(aes(S, ..density..)) +
    geom_histogram(color = "black", binwidth = 10) +
    ylab("Probability") +
    geom_line(data = normal_density, mapping = aes(s, f), color = "blue")

```

### Assessment: Questions 1 SAT testing

NA 

### Assessment: Questions 2 SAT testing

NA

### Assessment - Betting on Roulette
NA

## The Big Short: Interest Rates Explained
```{r}

#Code: Interest rate sampling model
n <- 1000
loss_per_foreclosure <- -200000
p <- 0.02
defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE)
sum(defaults * loss_per_foreclosure)

#Code: Interest rate Monte Carlo simulation
B <- 10000
losses <- replicate(B, {
    defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE) 
  sum(defaults * loss_per_foreclosure)
})

#Code: Plotting expected losses
library(tidyverse)
data.frame(losses_in_millions = losses/10^6) %>%
    ggplot(aes(losses_in_millions)) +
    geom_histogram(binwidth = 0.6, col = "black")

#Code: Expected value and standard error of the sum of 1,000 loans
n*(p*loss_per_foreclosure + (1-p)*0)    # expected value 
sqrt(n)*abs(loss_per_foreclosure)*sqrt(p*(1-p))    # standard error

#Code: Calculating interest rates for expected value of 0
x = - loss_per_foreclosure*p/(1-p)
x

#On a $180,000 loan, this equals an interest rate of:
x/180000

#Code: Calculating interest rate for 1% probability of losing money
l <- loss_per_foreclosure
z <- qnorm(0.01)
#x <- -l*( n*p - z*sqrt(n*p*(1-p)))/ ( n*(1-p) + z*sqrt(n*p*(1-p)))\x
x/180000    # interest rate
loss_per_foreclosure*p + x*(1-p)    # expected value of the profit per loan
n*(loss_per_foreclosure*p + x*(1-p)) # expected value of the profit over n loans

#Code: Monte Carlo simulation for 1% probability of losing money
B <- 100000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)    # expected value of the profit over n loans
mean(profit<0)    # probability of losing money

```
## The Big Short
```{r}
#Code: Expected value with higher default rate and interest rate
p <- .04
loss_per_foreclosure <- -200000
r <- 0.05
x <- r*180000
loss_per_foreclosure*p + x*(1-p)

#Code: Calculating number of loans for desired probability of losing money
#The number of loans required is:
z <- qnorm(0.01)
l <- loss_per_foreclosure
n <- ceiling((z^2*(x-l)^2*p*(1-p))/(l*p + x*(1-p))^2)
n    # number of loans required
n*(loss_per_foreclosure*p + x * (1-p))    # expected profit over n loans

#Code: Monte Carlo simulation with known default probability
B <- 10000
p <- 0.04
x <- 0.05 * 180000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)

#Code: Monte Carlo simulation with unknown default probability
p <- 0.04
x <- 0.05*180000
profit <- replicate(B, {
    new_p <- 0.04 + sample(seq(-0.01, 0.01, length = 100), 1)
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-new_p, new_p), replace = TRUE)
    sum(draws)
})
mean(profit)    # expected profit
mean(profit < 0)    # probability of losing money
mean(profit < -10000000)    # probability of losing over $10 million

```
### Assessment: Questions 1 and 2: Insurance rates, part 1
NA

### Assessment: Questions 3 and 4: Insurance rates, part 2
NA

