---
title: "Productivity Tools: Section 3"
author: "Gustavo Roa"
date: "2023-10-03"
format: html
editor: source
---

# Texbook

<https://rafalab.dfci.harvard.edu/dsbook/>

# Section 3

## Reproducible Reports with R Markdown

**Key points:** - The final output is usually a report, textual descriptions and figures, and tables. - The aim is to generate a reproducible report in R markdown and knitr. - Features of Rmarkdown: code and text can be combined to the same document and figures and tables are automatically added to the file.

## R Markdown

**Key points:** - R Markdown is a format for literate programming documents. Literate programming weaves instructions, documentation and detailed comments in between machine executable code, producing a document that describes the program that is best for human understanding.

-   Start an R markdown document by clicking on File \> New File \> the R Markdown

-   The output could be HTML, PDF, or Microsoft Word, which can be changed in the header output, e.g. pdf_document / html_document/p\>

Code

```{r}
summary(pressure)

# a sample code chunk

# When echo=FALSE, code will be hided in output file

#use a descriptive name for each chunk for debugging purpose

```

## knitr

**Key points:**

-   The knitr package is used to compile R markdown documents.
-   The first time you click the "knit" button on the editor pane a pop-up window will prompt you to install packages, but after that is completed,
-   the button will automatically knit your document.
-   github_document gives a .md file, which will give the best presentation on GitHub

Code

```{r}
# \# output: html_document \# output: pdf_document \# output: word_document \# output: github_document

```

## Comprehension Check: Reproducible Reports

### Question 1

Why might you want to create a report using R Markdown?

-   R Markdown has better spell-checking tools than other word processors.

-   R Markdown allows you to automatically add figures to the final document.

-   R Markdown final reports have smaller file sizes than Word documents.

-   R Markdown documents look identical to the final report.

### Question 2

You have a vector of student heights called heights. You want to generate a histogram of these heights in a final report, but you don't want the code to show up in the final report. You want to name the R chunk "histogram" so that you can easily find the chunk later.

Which of the following R chunks does everything you want it to do?

```{r}
# {r, histogram, message=FALSE} hist(heights)

# {r histogram, warning=FALSE} hist(heights)

# {r, echo=FALSE} hist(heights)

# {r histogram, echo=FALSE} hist(heights)
```

### Question 3

Below is a section of R Markdown code that generates a report.

title: "Final Grade Distribution" output: pdf_document

```{r}
# {r, echo=FALSE} load(file="my_data.Rmd") summary(grades)

```

Select the statement that describes the file report generated by the R markdown code above.

-   A PDF document called "Final Grade Distribution" that prints a summary of the "grades" object. The code to load the file and produce the summary will not be included in the final report.

-   A PDF document called "Final Grade Distribution" that prints a summary of the "grades" object. The code to load the file and produce the summary will be included in the final report.

-   An HTML document called "Final Grade Distribution" that prints a summary of the "grades" object. The code to load the file and produce the summary will not be included in the final report.

-   A PDF document called "Final Grade Distribution" that is empty because the argument echo=FALSE was used.

### Question 4

The user specifies the output file format of the final report when using R Markdown.

Which of the following file types is NOT an option for the final output?

-   .rmd

-   .pdf

-   .doc

-   .html

### Question

```{r}
# {r, echo=F} n \<- nrow(mtcars)
```

Here 'r n' cars are compared

What will be the output for the above Rmarkdown file when knit to HTML?

-   The only output is the text: Here 32 cars are compared.

-   Since we have echo=F, the code chunk is not evaluated, therefore we will have both the code and the text: Here 'r n' cars are compared.

-   The code will be displayed as well as Here 32 cars are compared.

-   R cannot comprehend the value of n, we will get an error.
